%!TEX root = manual.tex
%
\chapter{Introduction}
\label{chap:introduction}

\thispagestyle{empty}

The \emph{Abstract Syntax Description Lanuguage} (\asdl{}) is a language
designed to describe the tree-like data structures in compilers.
Its original purpose to provide a method for
compiler components written in different languages to
interoperate.
\asdl{} makes it easier for applications written in a
variety of programming languages to communicate complex recursive data
structures. 

\asdlgen{} is a tool that takes \asdl{} descriptions and produces
implementations of those descriptions in a variety of languages.
\asdl{} and \asdlgen{} together provide the following advantages
\begin{itemize}
  \item Concise descriptions of important data structures.
  \item Automatic generation of data structure implementations for
    \asdlgen{}-supported languages.
  \item Automatic generation of functions to read and write the data
    structures to disk in a machine and language independent way.
\end{itemize}%

\asdl{} descriptions describe the tree-like data structures such as
abstract syntax trees (ASTs) and compiler intermediate representations
(IRs).
Tools such as \asdlgen{} automatically produce the equivalent
data structure definitions for the supported languages.
\asdlgen{} also produces functions for each language that read and
write the data structures to and from a platform and language
independent sequence of bytes.
The sequence of bytes is called a \emph{pickle}.

\asdl{} was originally developed in the 1990's by Daniel Wang as part of the
\emph{National Compiler Infrastructure} project at Princeton University.
Tha \asdl{} implementation has not kept up with the significant changes
in many of its target languages (\eg{}, \Cplusplus{}, \haskell{}, \java{},
\etc{}), so it was time for a rewrite.\footnote{
  Version~2.0 of \asdl{} is still available from
  \url{https://sourceforge.net/projects/asdl}
  and the~2.0 version of the manual (converted to \LaTeX{} is included
  in the documentation of this system.
}
Version~3.0 of \asdl{} and \asdlgen{} is a complete reimplementation of the
system, with the primary purpose of supporting the \smlnj{} compiler.
As such, it currently only supports generating picklers in \sml{} and
modern \Cplusplus{}, although other languages may be added as time
permits.

\section{Changes From Version 2.0}
The following is a list of the major changes from the 2.0 version of \asdl{}
and \asdlgen{}:
\begin{itemize}
  \item
    Currently only two target languages are supported: \sml{} and \Cplusplus{}.
  \item
    The generated \Cplusplus{} code targets the 2011 standard and uses the
    STL (\eg{}, \lstinline[language=C++]!std::vector<>! for \asdl{} sequences).
  \item
    Data can be pickled/unpickled to/from memory, as well as files.
    This change affects the requirements for implementing primitive modules.
\end{itemize}%
